---
title: 'Final Data Project'
author: "STAT 420, Summer 2021, Warren Child, Zoheb Satta, Yoga Mahalingam"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
library(knitr)
opts_chunk$set(cache = TRUE, autodep = TRUE)
```

# Life Expectancy Model

## Proposal
This project will employ data on variables that affect life expectancy. The data--which originates from the World Health Organization and which is available on Kaggle (https://www.kaggle.com/kumarajarshi/life-expectancy-who)--consists of 2939 observations on 22 variables. Our response variable will be life expectancy, and we will attempt to develop a model that minimizes prediction errors for life expectancy. The data contains two categorical variables--country and level of economic development. The rest of the variables are numerical.

Many of the variables fall under discrete themes. For example, there are variables for diseases, variables concerning the economy and population, and variables related to nutrition. When developing our model, we must consider how to treat variables that fall under such themes. These theme-based variables may end up exhibiting significant collinearity, or they may prove to be legitimate independent variables. 

Another important consideration is that each record is associated with a year and a country. After discussion with the professor or a TA, we must decide whether to ignore country (as a factor it would create a huge number of parameters), as well as whether to average all of the years for a given country. One possibility is that, rather than using country as a factor variable, it may be more useful to bin countries into a number of economic levels based on per-capita GDP. 

## Datasource

https://www.kaggle.com/kumarajarshi/life-expectancy-who

## Data summary
```{r}
library(readr)
data = read_csv("Life Expectancy Data.csv")
```

```{r}
str(data)
names(data)
```

- Number of observations: ``r {nrow(data)}``
- Number of variables : ``r {ncol(data)}``
- Response variable: `Life expectancy` Type: `Numeric`
- Potential Categorical Predictor variables:
  - Country
  - Status
  - Year
- Potential Numeric Predictor varianles:
  -Adult Mortality
  - infant deaths
  - Alcohol
  - percentage expenditure
  - Hepatitis B
  - Measles
  - BMI
  - under-five deaths
  - Polio
  - Total expenditure
  - Diphtheria
  - HIV/AIDS
  - GDP
  - Population
  - thinness  1-19 years
  - thinness 5-9 years
  - Income composition of resources
  - Schooling

```{r}
head(data)
```
