---
title: 'Final Data Project'
author: "STAT 420, Summer 2021, Warren Daniel Child, Zoheb Satta, Yoga Mahalingam"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
library(knitr)
opts_chunk$set(cache = TRUE, autodep = TRUE)
```

# Life Expectancy Model

## Proposal

## Datasource

https://www.kaggle.com/kumarajarshi/life-expectancy-who

## Data summary
```{r}
library(readr)
data = read_csv("Life Expectancy Data.csv")
```

```{r}
str(data)
names(data)
```

- Number of observations: ``r {nrow(data)}``
- Number of variables : ``r {ncol(data)}``
- Response variable: `Life expectancy` Type: `Numeric`
- Potential Categorical Predictor variables:
  - Country
  - Status
  - Year
- Potential Numeric Predictor varianles:
  -Adult Mortality
  - infant deaths
  - Alcohol
  - percentage expenditure
  - Hepatitis B
  - Measles
  - BMI
  - under-five deaths
  - Polio
  - Total expenditure
  - Diphtheria
  - HIV/AIDS
  - GDP
  - Population
  - thinness  1-19 years
  - thinness 5-9 years
  - Income composition of resources
  - Schooling

```{r}
head(data)
```
